A preliminary constraint grammar for Russian

### 2015-05-04 2015-nodalida-ruscg/ruscg.tex
more fixes
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60226 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{document}
\maketitle
\begin{abstract}
 This paper presents preliminary work on a constraint
 grammar based disambiguator for Russian. Russian is
 a Slavic language with a high degree of both in-category
 and out-category homonymy in the inflectional system.
 The pipeline consists of a finite-state morphological
 analyser and constraint grammar. The constraint 
 grammar is tuned to be high recall at the expense of 
 low precision.
 \rr{add results to abstract}
\end{abstract}

@@

\begin{table}[hb]
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
  \label{tab:ruleDist}
\end{table}

@@

Due to the lack of a hand-annotated corpus compatible with our morphological
analyser, we adopted a slightly modified technique. We picked a random text
from the Russian Wikipedia, ran it through the morphological analyser, wrote
\rr{R1: Is ru-wikipedia representative of Russian texts on the whole?}
rules, and then ran the rules on the whole Wikipedia corpus. For each rule,
we collected around 100 example applications and checked them. If a rule
selected the appropriate reading in all cases, we included it in the \emph{safe}
rule set, if it removed an appropriate reading in less then three cases, 
then we included it in the \emph{safe heuristic} rule set. Otherwise we either
discarded the rule or included it in the heuristic rule set.

@@

\subsection{Corpus}
In order to evaluate the grammar we hand-annotated 10,150 words of Russian text
from Wikipedia articles, public domain literature and freely-available news sources. The
annotated texts are available online under the {\sc cc-by-sa} licence.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-rus/texts/}}\rr{is this a good url? Do we need to put a COPYING file in that folder?}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{graphics/learning-curve.pdf}
  \caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
    \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
    Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure}
\rr{R1: Figure 3 is small and hard to read. Shapes indistinguishable.}
\rr{R2: If I interpret Fig. 3 correctly, the HMM system alone is much more dependent on training corpus size than the voting set-up, where accuracies actually appear to become asymptotic already with 8-10,000 words in the training corpus, indicating that further improvement depends more on the CG than the HMM-component.}
\section{Future work}
\label{sec:future}
% ROB
% FRAN
% func analysis
% dep analysis

>>>>

\begin{document}
\maketitle
\begin{abstract}
 This paper presents preliminary work on a constraint
 grammar based disambiguator for Russian. Russian is
 a Slavic language with a high degree of both in-category
 and out-category homonymy in the inflectional system.
 The pipeline consists of a finite-state morphological
 analyser and constraint grammar. The constraint 
 grammar is tuned to be high recall (over 0.99) at the expense of 
 low precision.
% \rr{add results to abstract}
\end{abstract}

@@

\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
  \label{tab:ruleDist}
\end{table}

@@

Due to the lack of a hand-annotated corpus compatible with our morphological
analyser, we adopted a slightly modified technique. We picked a random text
from the Russian Wikipedia,\footnote{The Russian Wikipedia was chosen as a testing corpus
  as it is the largest, freely licensed corpus of Russian available on the internet. It is 
  not representative of Russian texts as a whole.}
ran it through the morphological analyser, wrote
%\rr{R1: Is ru-wikipedia representative of Russian texts on the whole?}
rules, and then ran the rules on the whole Wikipedia corpus. For each rule,
we collected around 100 example applications and checked them. If a rule
selected the appropriate reading in all cases, we included it in the \emph{safe}
rule set, if it removed an appropriate reading in less then three cases, 
then we included it in the \emph{safe heuristic} rule set. Otherwise we either
discarded the rule or included it in the heuristic rule set.

@@

\subsection{Corpus}
In order to evaluate the grammar we hand-annotated 10,150 words of Russian text
from Wikipedia articles, public domain literature and freely-available news sources. The
annotated texts are available online under the {\sc cc-by-sa} licence.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-rus/texts/}}
%\rr{is this a good url? Do we need to put a COPYING file in that folder?}

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.70\textwidth]{graphics/learning-curve.pdf}
  \caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
    \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
    Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure*}
\rr{R2: If I interpret Fig. 3 correctly, the HMM system alone is much more dependent on training corpus size than the voting set-up, where accuracies actually appear to become asymptotic already with 8-10,000 words in the training corpus, indicating that further improvement depends more on the CG than the HMM-component.}
\section{Future work}
\label{sec:future}
% ROB
% FRAN
% func analysis
% dep analysis
